---
title: "Class5: Data Viz with ggplot"
author: "Kaliyah Adei-Manu A18125684"
format: pdf
toc: true
---

## Background

There are lot's of ways to make figures in R.  These include so-caled "base R" graphics (e.g. `plot()`) and tons of add-on packages like **ggplot2**.

For example her we make the same plot with both : 

```{r}
head(cars)
```
```{r}
plot(cars)
```

First I need to install the package with the command `install.package()`

>**N.B** we never run an install cmd in a quarto code chunk or we will end up re-installing packages many times - which is not what we want!

Every time you want to use on of these "add-on" packages we need to load it up in R with the `library()` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things: 

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data maps to the plot
- The **geom**etry, the type of plot

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() 
```

Add a line to better show relationship between speed and distance 

```{r}
p <-  ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Stopping distance of old cars",
       subtitle = "Data from `cars` object",
       x="Speed (mph)",
       y="Distance (ft)")
```

```{r}
p + theme_bw()
```



## Gene expression plot

We can read the input 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot 

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point()
```


```{r}
table( genes$State )
```

Version 2 that is colored by `State` so we can see the "up" and "down" significant genes compared to all the "unchanging" genes

```{r}
nrow(genes)
colnames(genes)
ncol(genes)
```


```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```
Version 3 plot, let;s modify the default colors to something we like

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) + 
  geom_point()+
  scale_color_manual(values=c("blue","grey","red"))+
  labs(x="Control (no drug)", y= "Drug", 
       title = "Gene Expression Changes upon GLP-1 Drug ") +
  theme_bw()
```

## Going Further

Let's have a look at the famous **gapminder** dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```
```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above. 

```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col=continent) +
  geom_point(alpha=0.3)+
  facet_wrap(~continent)
```





## Custom Plots

How big is this gapminder data set?

```{r}
nrow(gapminder)
ncol(gapminder)
```
I want to "filter" down to a subset of this data.  I will use the **dplyr** package to help me. 

First I need to install it and then load it up...
`install.packages("dplyr")` and then `library(dplyr)`
```{r}
library(dplyr)

gapfinder_2007 <-filter(gapminder, year==2007)
head(gapfinder_2007)
```


```{r}
filter(gapfinder_2007, country == "Ireland")
```

```{r}
filter(gapminder, year == 2007, country == "Ireland")
```
```{r}
filter(gapminder, year == 1977, country == "United States")
```
```{r}
filter(gapminder, year == 1977, country == "Ireland")
```
```{r}
filter(gapminder, year == 2007, country == "United States")
```

> Q. Make a plt comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977, 2007)) 
ggplot(input) +
  aes(gdpPercap,lifeExp, col=continent) +
  geom_point() +
  facet_wrap(~year)
```

